---
title: "Exercise 8"
author: "Haoran Xu"
output:
  pdf_document: default
  rmdformats::robobook:
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r ch08-load-packages, message=FALSE, cache=FALSE, include=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
library(mlogit) # Multinomial Logit Models
library(gmnl) # Multinomial Logit Models with Random Parameters
library(gridExtra) # Miscellaneous Functions for "Grid" Graphics
library(stargazer) # Well-Formatted Regression and Summary Statistics Tables
library(tibble) # Simple Data Frames
library(tidyr) # Tidy Messy Data
```

## 1. Repeat the illustrative accept-reject simulation multiple times (e.g., `s` = 1000) using `r` = 100 and summarize the results in the form of a histogram or frequency polygon. Next, do the same experiment butwith `r` = 1000 and `r` = 5000. Compare to the exact area and discuss your results. Hint: You can simultaneously draw `s` × `r` random numbers and then collect them in `s` groups (see package {dplyr} functions `group_by()` and `summarize()`).

```{r s1000-and-r100}
# Set seed for replicability (this defines a starting number from which the sequence of random numbers is generated )
set.seed(35739)

norm_pdf <- data.frame(x = seq(0.2, 
                               1.2, 
                               0.01)) %>%
  mutate(f = dnorm(x, 
                   mean = 0, 
                   sd = 0.5))

# setting parameters
r <- 100
s <- 1000   # Number of simulations
xmin <- 0.2
xmax <- 1.2
fmax <- max(norm_pdf$f)  # Max height of PDF

sim_pdf <- data.frame(
  sim_group = rep(1:s, each = r),  # Create 1000 groups of 100 draws
  x = runif(s * r, xmin, xmax),     # All x-values
  fsim = runif(s * r, 0, fmax)      # All y-values
)

sim_pdf <- sim_pdf %>%
  mutate(
    f = dnorm(x, mean = 0, sd = 0.5),
    status = ifelse(fsim <= f, "accept", "reject")
  )

# Calculate acceptance rate per simulation group
accept_rates <- sim_pdf %>%
  group_by(sim_group) %>%
  summarize(accept_rate = sum(status == "accept") / r)

# Theoretical acceptance probability (for comparison)
theoretical_prob <- integrate(
  function(x) dnorm(x, 0, 0.5), 
  lower = xmin, 
  upper = xmax
)$value / ((xmax - xmin) * fmax)

# Plot histogram of acceptance rates
ggplot(accept_rates, aes(x = accept_rate)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    fill = "skyblue",
    color = "black"
  ) +
  geom_vline(
    xintercept = theoretical_prob,
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Acceptance Rate Distribution (1000 Simulations with 100 draws)",
    x = "Acceptance Rate",
    y = "Density"
  ) +
  annotate(
    "text",
    x = theoretical_prob + 0.02,
    y = 15,
    label = paste0("Theoretical: ", round(theoretical_prob, 3)),
    color = "red"
  )
```

The codes for \( r \) = 1000 and \( r \) = 5000 are similar, the histograms for each are listed below.

```{r s1000-and-r1000, echo=FALSE}
# Set seed for replicability (this defines a starting number from which the sequence of random numbers is generated )
set.seed(35739)

# setting parameters
r <- 1000
s <- 1000   # Number of simulations
xmin <- 0.2
xmax <- 1.2
fmax <- max(norm_pdf$f)  # Max height of PDF

sim_pdf2 <- data.frame(
  sim_group = rep(1:s, each = r),  # Create 1000 groups of 100 draws
  x = runif(s * r, xmin, xmax),     # All x-values
  fsim = runif(s * r, 0, fmax)      # All y-values
)

sim_pdf2 <- sim_pdf2 %>%
  mutate(
    f = dnorm(x, mean = 0, sd = 0.5),
    status = ifelse(fsim <= f, "accept", "reject")
  )

# Calculate acceptance rate per simulation group
accept_rates2 <- sim_pdf2 %>%
  group_by(sim_group) %>%
  summarize(accept_rate = sum(status == "accept") / r)

# Plot histogram of acceptance rates
ggplot(accept_rates2, aes(x = accept_rate)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    fill = "skyblue",
    color = "black"
  ) +
  geom_vline(
    xintercept = theoretical_prob,
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Acceptance Rate Distribution (1000 Simulations with 1000 draws)",
    x = "Acceptance Rate",
    y = "Density"
  ) +
  annotate(
    "text",
    x = theoretical_prob + 0.02,
    y = 15,
    label = paste0("Theoretical: ", round(theoretical_prob, 3)),
    color = "red"
  )
```

```{r s1000-and-r5000, echo=FALSE}
# Set seed for replicability (this defines a starting number from which the sequence of random numbers is generated )
set.seed(35739)

# setting parameters
r <- 5000
s <- 1000   # Number of simulations
xmin <- 0.2
xmax <- 1.2
fmax <- max(norm_pdf$f)  # Max height of PDF

sim_pdf3 <- data.frame(
  sim_group = rep(1:s, each = r),  # Create 1000 groups of 100 draws
  x = runif(s * r, xmin, xmax),     # All x-values
  fsim = runif(s * r, 0, fmax)      # All y-values
)

sim_pdf3 <- sim_pdf3 %>%
  mutate(
    f = dnorm(x, mean = 0, sd = 0.5),
    status = ifelse(fsim <= f, "accept", "reject")
  )

# Calculate acceptance rate per simulation group
accept_rates3 <- sim_pdf3 %>%
  group_by(sim_group) %>%
  summarize(accept_rate = sum(status == "accept") / r)

# Plot histogram of acceptance rates
ggplot(accept_rates3, aes(x = accept_rate)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    fill = "skyblue",
    color = "black"
  ) +
  geom_vline(
    xintercept = theoretical_prob,
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Acceptance Rate Distribution (1000 Simulations with 5000 draws)",
    x = "Acceptance Rate",
    y = "Density"
  ) +
  annotate(
    "text",
    x = theoretical_prob + 0.02,
    y = 15,
    label = paste0("Theoretical: ", round(theoretical_prob, 3)),
    color = "red"
  )
```

The simulations demonstrate that increasing the number of draws per trial (\(r\)) significantly reduces variability in acceptance rate estimates. For \(r = 100\), the histogram of acceptance rates across \(s = 1000\) trials shows wider dispersion around the theoretical value than the other two trials. With \(r = 1000\), the distribution narrows more, and for \(r = 5000\), the histogram becomes more like normal distribution, being symmetric on the both sides theretical value line. This illustrates the **Law of Large Numbers** theory, which means that larger \(r\) yields more precise approximations of the true probability.


## 2. Re-estimate the multinomial probit model in this chapter, still with `r` = 50 but using a different random seed. Do this several times. The estimated values of the parameters will change, which is to be expected...but which ones change more? Why do you think this happens?

```{r ch08-load-data, include=FALSE}
data("Heating")
```

```{r ch08-prepare-data, include=FALSE}
H <- mlogit.data(Heating, 
                 shape = "wide", 
                 choice = "depvar", 
                 varying = c(3:12))
```

```{r, results="asis", cache=TRUE}
prob1 <- mlogit(depvar ~ ic + oc, 
                H, 
                reflevel = "gc", 
                probit = TRUE, 
                # can set the seed to control the random numbers (parameter1)
                seed = 3333,
                # number of draws (parameter2)
                R = 50)

prob2 <- mlogit(depvar ~ ic + oc, 
                H, 
                reflevel = "gc", 
                probit = TRUE, 
                # can set the seed to control the random numbers (parameter1)
                seed = 1069,
                # number of draws (parameter2)
                R = 50)

prob3 <- mlogit(depvar ~ ic + oc, 
                H, 
                reflevel = "gc", 
                probit = TRUE, 
                # can set the seed to control the random numbers (parameter1)
                seed = 2024,
                # number of draws (parameter2)
                R = 50)
```

```{r, echo=FALSE}
prob1.summary <- rownames_to_column(data.frame(summary(prob1)$CoefTable),
                                     "Variable") %>%
  transmute(Variable,
            Estimate, 
            pval = `Pr...z..`)

prob2.summary <- rownames_to_column(data.frame(summary(prob2)$CoefTable), 
                                     "Variable") %>%
  transmute(Variable,
            Estimate,
            pval = `Pr...z..`)

prob3.summary <- rownames_to_column(data.frame(summary(prob3)$CoefTable),
                                   "Variable") %>% 
  transmute(Variable, 
            Estimate, 
            pval = `Pr...z..`)

prob_table_1 <- full_join(prob1.summary, 
                          prob2.summary, 
                          by = "Variable") %>% 
  full_join(prob3.summary,
            by = "Variable")

kable(prob_table_1, 
      "latex",
      digits = 4,
      col.names = c("Variable",
                    "Estimate",
                    "p-value",
                    "Estimate",
                    "p-value",
                    "Estimate",
                    "p-value"),
      caption = "\\label{tab:table-mixed-probit-models}Multinomial Probit models with different seeds",
      ) %>%
  kable_styling() %>%
  add_header_above(c(" " = 1, "Seed = 3333" = 2, "Seed = 1069" = 2, "Seed = 2024" = 2)) %>%
  footnote(general = c(paste0("Log-Lik (seed=3333):",
                              round(prob1$logLik, 
                                    digits = 3),
                            "; Log-Lik (seed=1069):", 
                            round(prob2$logLik, 
                                  digits = 3),
                            "; Log-Lik (seed=2024):", 
                            round(prob3$logLik, 
                                  digits = 3))))
```

Re-estimating the multinomial probit model with different seeds (`seed = 3333`, `1069`, `2024`) and `R = 50` reveals different patterns in parameter stability. First, The main coefficients (i.e., `ic` and `oc`) remain remarkably consistent across different runs. In contrast, intercepts and covariance parameters exhibit significant variability. For example, the intercept of `(Intercept):gr` ranges from -48.69 to -21.51, with uniformly insignificant p-values < 0.68, indicating poor identification. Meanwhile, covariance terms like `ec.gr` (9.26 to 3.27) and `ec.er` (-0.01 to 0.07) vary widely, with p-values > 0.9.

The reasons for can be concluded in several aspects: a) Limited draws. The Halton sequence approximation for the multivariate normal integrals becomes noisy with fewer draws, disproportionately affecting covariance terms and intercepts tied to latent utilities. b) Identification challenges. Covariance parameters and intercepts for rarely chosen alternatives (e.g., `gr`, `hp`) are poorly identified, amplifying sensitivity to simulation noise. One way to improve is to increase `R` for stable covariance estimation or simplify the model by restricting off-diagonal terms. This underscores the trade-off between computational cost and precision in multinomial probit models.


## 3. Use the models you estimated in Question 2 and repeat the scenario where the installation cost of gas central systems increases by 15%. What can you say about the patterns of substitution when you calculate them with different models? What would you do to be more confident about the estimated patterns of substitution?

```{r, results="asis", cache=TRUE, echo=FALSE}
H_increase <- H
H_increase[H_increase$alt == "gc", "ic"] <- 1.15 * H_increase[H_increase$alt == "gc", "ic"]

scenario_prob1 = data.frame(Policy = c("Do nothing", "15% increase (seed = 3333)"),
                         rbind(apply(predict(prob1, 
                                             newdata = H), 
                                     2, 
                                     mean),
                               apply(predict(prob1, 
                                             newdata = H_increase),
                                     2, 
                                     mean)))

# Render as LaTeX table with booktabs formatting
knitr::kable(
  scenario_prob1,
  format = "latex",
  caption = "Predicted Choice Probabilities Under Different Policies (seed = 3333)",
  booktabs = TRUE,
  align = c("l", "r", "r", "r", "r"), # Align columns (left for Policy, right for others)
  digits = 3 # Round numeric columns to 3 decimal places
) %>%
  kableExtra::kable_styling(latex_options = "hold_position")

scenario_prob2 = data.frame(Policy = c("Do nothing", "15% increase (seed = 1069)"),
                         rbind(apply(predict(prob2, 
                                             newdata = H), 
                                     2, 
                                     mean),
                               apply(predict(prob2, 
                                             newdata = H_increase),
                                     2, 
                                     mean)))

# Render as LaTeX table with booktabs formatting
knitr::kable(
  scenario_prob2,
  format = "latex",
  caption = "Predicted Choice Probabilities Under Different Policies (seed = 1069)",
  booktabs = TRUE,
  align = c("l", "r", "r", "r", "r"), # Align columns (left for Policy, right for others)
  digits = 3 # Round numeric columns to 3 decimal places
) %>%
  kableExtra::kable_styling(latex_options = "hold_position")

scenario_prob3 = data.frame(Policy = c("Do nothing", "15% increase (seed = 2024)"),
                         rbind(apply(predict(prob3, 
                                             newdata = H), 
                                     2, 
                                     mean),
                               apply(predict(prob3, 
                                             newdata = H_increase),
                                     2, 
                                     mean)))

# Render as LaTeX table with booktabs formatting
knitr::kable(
  scenario_prob3,
  format = "latex",
  caption = "Predicted Choice Probabilities Under Different Policies (seed = 2024)",
  booktabs = TRUE,
  align = c("l", "r", "r", "r", "r"), # Align columns (left for Policy, right for others)
  digits = 3 # Round numeric columns to 3 decimal places
) %>%
  kableExtra::kable_styling(latex_options = "hold_position")


Ratios_prob <- data.frame(rbind(
  (scenario_prob1[2,2:6] - scenario_prob1[1,2:6]) / scenario_prob1[1,2:6] * 100,
  (scenario_prob2[2,2:6] - scenario_prob2[1,2:6]) / scenario_prob1[1,2:6] * 100,
  (scenario_prob3[2,2:6] - scenario_prob3[1,2:6]) / scenario_prob1[1,2:6] * 100
))

row.names(Ratios_prob) <- c("Ratio of probabilities (seed = 3333)",
                            "Ratio of probabilities (seed = 1069)",
                            "Ratio of probabilities (seed = 2024)") 

# Render as LaTeX table with booktabs formatting
knitr::kable(
  Ratios_prob,
  format = "latex",
  caption = "Substitution Effects between different models with different seeds",
  booktabs = TRUE,
  align = c("l", "r", "r", "r", "r"), # Align columns (left for Policy, right for others)
  digits = 3 # Round numeric columns to 3 decimal places
) %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

The substitution effects across models with different seeds (`3333`, `1069`, `2024`) reveal consistent trends but notable variability in magnitude. A 15% increase in gas central (`gc`) installation costs drives significant substitution toward electric central (`ec`) and electric room (`er`) systems, with their probability ratios increasing by ~30–33% and ~26–27%, respectively. However, substitutions to gas room (`gr`) and heat pump (`hp`) systems are smaller and more erratic, with `gr` fluctuating between 0.2–0.85%, and `hp` varying between 23–26%.

To improcec onfidence in the estimates, two ways could be implemented: a) Increase simulation draws. Raising `R` from 50 to more than 1,000 would reduce noise in the Halton sequence approximation, stabilizing covariance estimates and substitution patterns, though this would come with bigger computational costs. b) Robustness checks. Replicate the analysis across more seeds and models to distinguish true substitution trends from simulation artifacts. 

## 4. Re-estimate the multinomial probit model in this chapter with `r` = 50 and the seed set to 3245, but change the reference level to “ec”. What happens? What do you think are the implications for experimenting with multinomial probit model specifications?

```{r, results="asis", cache=TRUE, eval=FALSE}
prob4 <- mlogit(depvar ~ ic + oc, 
                H,
                reflevel = "ec", 
                probit = TRUE, 
                # can set the seed to control the random numbers (parameter1)
                seed = 3245,
                # number of draws (parameter2)
                R = 50)
```

Changing the reference level to “ec” (electric central) triggers a computational error regarding the "Cholesky decomposition of the covariance matrix". The error `the leading minor of order 1 is not positive` indicates that the covariance matrix of the latent utilities is not positive definite when “ec” is the base category. This likely arises because the model’s covariance structure becomes near-singular due to collinearity between alternatives.

The implications are that when working with multinomial probit models, the choice of reference level can directly affect model results. Certain reference categories may introduce structural dependencies in the error covariance matrix, leading to estimation failures, therefore, experimentation with differnt reference levels is encouraged.

