---
title: "Exercise5"
author: "Haoran Xu"
date: "2025-02-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, message=FALSE, cache=FALSE}
library(discrtr) # A companion package for the book Introduction to Discrete Choice Analysis with `R` 
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(htmlwidgets) # HTML Widgets for R
library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax 
library(mlogit) # Multinomial Logit Models 
library(plotly) # Create Interactive Web Graphics via 'plotly.js' 
library(stargazer) # Well-Formatted Regression and Summary Statistics Tables 
library(tidyr) # Tidy Messy Data
```

```{r load-and-rename-data}
data("mc_commute_wide",
     package = "discrtr") 
```


```{r reshape-full-table}
mc_commute_long <- mc_commute_wide %>%
  mlogit.data(shape="wide", 
              # Name of column with the choices
              choice = "choice", 
              # Numbers of columns with attributes that vary by alternative
              varying = 3:22)
```

```{r formula-1}
# Function `mFormula()` is used to define multi-part formulas of the form:
# y ~ x | z | w, which in the notation used for the anatomy of utility functions is
# choice ~ alternative vars. with generic coefficients |
#          individual vars. with specific coefficients |
#          alternative vars. with specific coefficients 
# In this formula time is one of x variables
f1 <- mFormula(choice ~ time)
```

```{r formula-2}
# Function `mFormula()` is used to define multi-part formulas of the form:
# y ~ x | z | w, which in the notation used for the anatomy of utility functions is
# choice ~ alternative vars. with generic coefficients |
#          individual vars. with specific coefficients |
#          alternative vars. with specific coefficients 
# In this formula `time` is one of x variables and `sidewalk_density` is one of z variables
f2 <- mFormula(choice ~ time | sidewalk_density)
```

```{r formula-3}
f3 <- mFormula(choice ~ 0 | sidewalk_density | time)
```

```{r summary-model-1}
# Function `mlogit()` is used to estimate logit models
# It needs a multi-part formula and a data set in long form
model1 <- mlogit(f1,
                 mc_commute_long)

# Function `summary()` give the summary of data objects,
# including the output of model estimation algorithms
summary(model1)
```

```{r summary-model-1-stargazer, results="asis"}
# Note: use chunk option results="asis" to display latex output in pdf
stargazer::stargazer(model1, 
                     # Use type = "text", "latex". or "html" depending
                     # on the desired output
                     type ="text",
                     header = FALSE,
                     single.row = TRUE,
                     title = "Estimation results: Model 1")
```

```{r summary-model-2-stargazer, results="asis"}
# Note: use chunk option results="asis" to display latex output in pdf
model2 <- mlogit(f2, 
                 mc_commute_long)

stargazer::stargazer(model2,
                     # Use type = "text", "latex". or "html" depending
                     # on the desired output
                     type = "text",
                     header = FALSE,
                     single.row = TRUE,
                     title = "Estimation results: Model 2")
```

## 1. In the example in this chapter we estimated the probabilities of choosing different
modes by sidewalk density setting travel time to the in-sample median. Use
`model2` to calculate the probability of choosing different modes by in-sample
median sidewalk density but now for travel times of 20, 30, and 40 min. Discuss
the results.

```{r create-matrix-to-predict, include=FALSE}
mc_commute_predict <- mc_commute_long[1:12,]
```

```{r set-values-for-predictions, include=FALSE}
# Function `rep()` repeats the values in the argument a designated
# number of times; here, the values in the sequence 0 to 60 in intervals
# of 5 are repeated four times each (once for each alternative)
mc_commute_predict$time <- rep(seq(from = 20,
                                   to = 40,
                                   by = 10),
                                   each = 4)
```

```{r median-commute-time, include=FALSE}
median(mc_commute_predict$sidewalk_density, 
       na.rm = TRUE)
```

```{r set-time-for-prediction, include=FALSE}
mc_commute_predict$time <- 22.63
```

```{r predict-probabilities-model-2, include=FALSE}
probs <- predict(model2,
                 newdata = mc_commute_predict)
```

```{r reshape-table, include=FALSE}
probs <- data.frame(time = seq(from = 20,
                               to = 40,
                               by = 10),
                    probs) %>% 
  # Pivot longer all columns _except_ `sidewalk_density`
  pivot_longer(cols = -time,
               # The column names become a new column called "Mode"
               names_to = "Mode",
               # The values are gathered into a single column called
               # "Probability"
               values_to = "Probability")
```

```{r probability-plot, echo=FALSE}
probs %>%
  # Create ggplot object; map `sidewalk_density` to the y-axis
  # `Probability` to the x-axis, and the color of geometric
  # objects to `Mode`
  ggplot(aes(x = time, 
             y = Probability,
             color = Mode)) +
  # Add geometric object of type line with size = 1
  geom_line(size = 1) +
  labs(y="Probability", 
       x = expression("Time (min)")) 
```


## 2. Estimate a model using formula `f3` (call it `model3`). Discuss the output of this
model.

```{r summary-model-3-change-of-reference, results="asis"}
# Note: use chunk option results="asis" to display latex output in pdf
model3 <- mlogit(f3,
                 mc_commute_long, 
                 # Specify the alternative that acts as reference
                 reflevel = "Walk")

stargazer::stargazer(model3, 
                     # Use type = "text", "latex". or "html" depending
                     # on the desired output
                     type = "text",
                     header = FALSE,
                     single.row = TRUE,
                     title = "Estimation results: Model 3 (alternative: Walk)")
```


## 3. Use `model3` to calculate the probability of choosing different modes by insample
median sidewalk density but now for travel times of 20, 30, and 40 min.
Discuss the results.

```{r predict-probabilities-model-3}
probs3 <- predict(model3,
                 newdata = mc_commute_predict)
```

```{r}
probs3 <- data.frame(time = seq(from = 20,
                               to = 40,
                               by = 10),
                    probs3) %>% 
  # Pivot longer all columns _except_ `sidewalk_density`
  pivot_longer(cols = -time,
               # The column names become a new column called "Mode"
               names_to = "Mode",
               # The values are gathered into a single column called
               # "Probability"
               values_to = "Probability")
```

```{r}
probs3 %>%
  # Create ggplot object; map `sidewalk_density` to the y-axis
  # `Probability` to the x-axis, and the color of geometric
  # objects to `Mode`
  ggplot(aes(x = time, 
             y = Probability,
             color = Mode)) +
  # Add geometric object of type line with size = 1
  geom_line(size = 1) +
  labs(y="Probability", 
       x = expression("Time (min)")) 
```


## 4. In the general case, what is the value of the log-likelihood of the null (Equally
Likely) model?

\[
LL_{null} = -N * log{J}
\]

## 5. Use the likelihood ratio test to compare `model3` to `model2`? Discuss the results.
What restrictions would you need to impose in `model3` to obtain `model2`?

```{r likelihood-ratio-test-2-3}
lrtest(model2,
       model3)
```

The very small p-value (< 2.2e-16) means we strongly reject the null hypothesis, which suggests that Model 2 fits significantly better than Model 1. Since Model 2 has a better fit, this indicates that including "time" as an alternative-specific predictor provides valuable information for modeling "choice".

Restrictions: setting the alternative var. with specific coefficients (\(\beta_2\), \(\beta_3\), \(\beta_4)\) of `model3` to be the same value of \(\beta_1\).

