---
title: "Exercise10"
author: "Haoran Xu"
output:
  rmdformats::robobook:
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r ch10-load-packages, message=FALSE, cache=FALSE}
library(AER) # Applied Econometrics with R 
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(gmnl) # Multinomial Logit Models with Random Parameters
library(gridExtra) # Miscellaneous Functions for "Grid" Graphics
library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
library(mlogit) # Multinomial Logit Models
library(stargazer) # Well-Formatted Regression and Summary Statistics Tables
library(tibble) # Simple Data Frames
library(tidyr) # Tidy Messy Data
```

```{r ch10-load-data} 
data("TravelMode", 
     package = "AER")
```


## 1. What is the difference between an error component and a random coefficient in
a mixed logit model?



## 2. In your own words, explain the IIA property as it pertains to the mixed logit
model.



## 3. Load the following data set from the `mlogit` package. This data set includes information about travel by mode in the Montreal–Toronto corridor. Variable `noalt` is the number of alternatives available to each respondent. Filter all respondents with four alternatives and process the data for use in `gmnl`. Estimate a mixed logit model using this data set. Justify your choice of variables. Which variables do you choose to have random coefficients and why?

```{r}
data("ModeCanada",
     package = "mlogit")
```

```{r}
MC <- mlogit.data(ModeCanada %>% 
                    filter(noalt == 4), 
                  choice = "choice", 
                  shape = "long",
                  alt.levels = c("air", 
                                 "train", 
                                 "bus", 
                                 "car"))
```


## 4. Graphically show the random coefficient(s) of your model, both the unconditional and conditional distributions. Discuss the results.





