---
title: "Exercise8"
author: "Haoran Xu"
output:
  rmdformats::robobook:
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r ch08-load-packages, message=FALSE, cache=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
library(mlogit) # Multinomial Logit Models
```

```{r ch08-load-data}
data("Heating")
```

```{r ch08-prepare-data}
H <- mlogit.data(Heating, 
                 shape = "wide", 
                 choice = "depvar", 
                 varying = c(3:12))
```

## 1. Repeat the illustrative accept-reject simulation multiple times (e.g., `s` = 1000) using `r` = 100 and summarize the results in the form of a histogram or frequency polygon.Next, do the same experiment butwith `r` = 1000 and `r` = 5000. Compare to the exact area and discuss your results. Hint: You can simultaneously draw `s` × `r` random numbers and then collect them in `s` groups (see package {dplyr} functions `group_by()` and `summarize()`).



## 2. Re-estimate the multinomial probit model in this chapter, still with `r` = 50 but using a different random seed. Do this several times. The estimated values of the parameters will change, which is to be expected...but which ones change more? Why do you think this happens?





## 3. Use the models you estimated in Question 2 and repeat the scenario where the installation cost of gas central systems increases by 15%.What can you say about the patterns of substitution when you calculate them with different models? What would you do to be more confident about the estimated patterns of substitution?




## 4. Re-estimate the multinomial probit model in this chapter with `r` = 50 and the seed set to 3245, but change the reference level to “ec”. What happens? What do you think are the implications for experimenting with multinomial probit model specifications?



