---
title: "Exercise 8"
author: "Haoran Xu"
output:
  pdf_document: default
  rmdformats::robobook:
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r ch08-load-packages, message=FALSE, cache=FALSE, include=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
library(mlogit) # Multinomial Logit Models
```

## 1. Repeat the illustrative accept-reject simulation multiple times (e.g., `s` = 1000) using `r` = 100 and summarize the results in the form of a histogram or frequency polygon. Next, do the same experiment butwith `r` = 1000 and `r` = 5000. Compare to the exact area and discuss your results. Hint: You can simultaneously draw `s` × `r` random numbers and then collect them in `s` groups (see package {dplyr} functions `group_by()` and `summarize()`).

```{r s1000-and-r100}
# Set seed for replicability (this defines a starting number from which the sequence of random numbers is generated )
set.seed(35739)

norm_pdf <- data.frame(x = seq(0.2, 
                               1.2, 
                               0.01)) %>%
  mutate(f = dnorm(x, 
                   mean = 0, 
                   sd = 0.5))

# setting parameters
r <- 100
s <- 1000   # Number of simulations
xmin <- 0.2
xmax <- 1.2
fmax <- max(norm_pdf$f)  # Max height of PDF

sim_pdf <- data.frame(
  sim_group = rep(1:s, each = r),  # Create 1000 groups of 100 draws
  x = runif(s * r, xmin, xmax),     # All x-values
  fsim = runif(s * r, 0, fmax)      # All y-values
)

sim_pdf <- sim_pdf %>%
  mutate(
    f = dnorm(x, mean = 0, sd = 0.5),
    status = ifelse(fsim <= f, "accept", "reject")
  )

# Calculate acceptance rate per simulation group
accept_rates <- sim_pdf %>%
  group_by(sim_group) %>%
  summarize(accept_rate = sum(status == "accept") / r)

# Theoretical acceptance probability (for comparison)
theoretical_prob <- integrate(
  function(x) dnorm(x, 0, 0.5), 
  lower = xmin, 
  upper = xmax
)$value / ((xmax - xmin) * fmax)

# Plot histogram of acceptance rates
ggplot(accept_rates, aes(x = accept_rate)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    fill = "skyblue",
    color = "black"
  ) +
  geom_vline(
    xintercept = theoretical_prob,
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Acceptance Rate Distribution (1000 Simulations with 100 draws)",
    x = "Acceptance Rate",
    y = "Density"
  ) +
  annotate(
    "text",
    x = theoretical_prob + 0.02,
    y = 15,
    label = paste0("Theoretical: ", round(theoretical_prob, 3)),
    color = "red"
  )
```

The codes for \( r \) = 1000 and \( r \) = 5000 are similar, the histograms for each are listed below.

```{r s1000-and-r1000, echo=FALSE}
# Set seed for replicability (this defines a starting number from which the sequence of random numbers is generated )
set.seed(35739)

# setting parameters
r <- 1000
s <- 1000   # Number of simulations
xmin <- 0.2
xmax <- 1.2
fmax <- max(norm_pdf$f)  # Max height of PDF

sim_pdf2 <- data.frame(
  sim_group = rep(1:s, each = r),  # Create 1000 groups of 100 draws
  x = runif(s * r, xmin, xmax),     # All x-values
  fsim = runif(s * r, 0, fmax)      # All y-values
)

sim_pdf2 <- sim_pdf2 %>%
  mutate(
    f = dnorm(x, mean = 0, sd = 0.5),
    status = ifelse(fsim <= f, "accept", "reject")
  )

# Calculate acceptance rate per simulation group
accept_rates2 <- sim_pdf2 %>%
  group_by(sim_group) %>%
  summarize(accept_rate = sum(status == "accept") / r)

# Plot histogram of acceptance rates
ggplot(accept_rates2, aes(x = accept_rate)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    fill = "skyblue",
    color = "black"
  ) +
  geom_vline(
    xintercept = theoretical_prob,
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Acceptance Rate Distribution (1000 Simulations with 1000 draws)",
    x = "Acceptance Rate",
    y = "Density"
  ) +
  annotate(
    "text",
    x = theoretical_prob + 0.02,
    y = 15,
    label = paste0("Theoretical: ", round(theoretical_prob, 3)),
    color = "red"
  )
```

```{r s1000-and-r5000, echo=FALSE}
# Set seed for replicability (this defines a starting number from which the sequence of random numbers is generated )
set.seed(35739)

# setting parameters
r <- 5000
s <- 1000   # Number of simulations
xmin <- 0.2
xmax <- 1.2
fmax <- max(norm_pdf$f)  # Max height of PDF

sim_pdf3 <- data.frame(
  sim_group = rep(1:s, each = r),  # Create 1000 groups of 100 draws
  x = runif(s * r, xmin, xmax),     # All x-values
  fsim = runif(s * r, 0, fmax)      # All y-values
)

sim_pdf3 <- sim_pdf3 %>%
  mutate(
    f = dnorm(x, mean = 0, sd = 0.5),
    status = ifelse(fsim <= f, "accept", "reject")
  )

# Calculate acceptance rate per simulation group
accept_rates3 <- sim_pdf3 %>%
  group_by(sim_group) %>%
  summarize(accept_rate = sum(status == "accept") / r)

# Plot histogram of acceptance rates
ggplot(accept_rates3, aes(x = accept_rate)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 30,
    fill = "skyblue",
    color = "black"
  ) +
  geom_vline(
    xintercept = theoretical_prob,
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Acceptance Rate Distribution (1000 Simulations with 5000 draws)",
    x = "Acceptance Rate",
    y = "Density"
  ) +
  annotate(
    "text",
    x = theoretical_prob + 0.02,
    y = 15,
    label = paste0("Theoretical: ", round(theoretical_prob, 3)),
    color = "red"
  )
```

The simulations demonstrate that increasing the number of draws per trial (\(r\)) significantly reduces variability in acceptance rate estimates. For \(r = 100\), the histogram of acceptance rates across \(s = 1000\) trials shows wider dispersion around the theoretical value than the other two trials. With \(r = 1000\), the distribution narrows more, and for \(r = 5000\), the histogram becomes more like normal distribution, being symmetric on the both sides theretical value line. This illustrates the **Law of Large Numbers** theory, which means that larger \(r\) yields more precise approximations of the true probability.


## 2. Re-estimate the multinomial probit model in this chapter, still with `r` = 50 but using a different random seed. Do this several times. The estimated values of the parameters will change, which is to be expected...but which ones change more? Why do you think this happens?

```{r ch08-load-data, include=FALSE}
data("Heating")
```

```{r ch08-prepare-data, include=FALSE}
H <- mlogit.data(Heating, 
                 shape = "wide", 
                 choice = "depvar", 
                 varying = c(3:12))
```





## 3. Use the models you estimated in Question 2 and repeat the scenario where the installation cost of gas central systems increases by 15%.What can you say about the patterns of substitution when you calculate them with different models? What would you do to be more confident about the estimated patterns of substitution?




## 4. Re-estimate the multinomial probit model in this chapter with `r` = 50 and the seed set to 3245, but change the reference level to “ec”. What happens? What do you think are the implications for experimenting with multinomial probit model specifications?



